%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% dissertation.prelim.tex
%
% Copyright 2012, Jeffrey Hellrung.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{On Embedded Methods for Crack Propagation, Virtual Surgery, Shattered Objects in Computer Animation, and Elliptic Partial Differential Equations}
\author{Jeffrey Lee Hellrung, Jr.}
\department{Mathematics}
\degreeyear{2012}

% Committee
\chair{Joseph M.\ Teran}
\member{Chris Anderson}
\member{Stanley J.\ Osher}
\member{Demetri Terzopoulos}

%\dedication{}

\acknowledgments{
I wish to thank the following for their support over the last several years as I've progressed through my doctoral degree here at UCLA:

\begin{itemize}
\item My family, especially my mom. I know it was difficult for her to resist the temptation to ask me about my degree progress every week, especially in the later years.
\item My girlfriend, Sirian; and her mom, Lisa.
\item The rest of my entering class who've remained in the department as long as I have. Or longer.
\end{itemize}

The research in Chapters~\ref{chap:partI.crackpropagation}, \ref{chap:partII.poisson}, and \ref{chap:partII.LE} was partially supported by UC Lab Fees Research / Department of Energy grant 09-LR-04-116741-BERA; Office of Naval Research grants N00014-03-1-0071 and N00014-10-1-0730; and National Science Foundation grant CCF-0830554. Additionally, the research in Chapter~\ref{chap:partI.crackpropagation} was partially supported by National Science Foundation grant DMS-0914813; and the research in Chapters~\ref{chap:partII.poisson} and \ref{chap:partII.LE} by National Science Foundation grants DMS-0502315 and DMS-0652427.

Chapter~\ref{chap:partI.crackpropagation} is a version (with moderate revisions) of ``An XFEM method for modeling geometrically elaborate crack propagation in brittle materials'' by Casey L. Richardson, Jan Hegemann, Eftychios Sifakis, Jeffrey Hellrung, and Joseph M. Teran (PI) in \textit{International Journal for Numerical Methods in Engineering} (88(10):1042--1065, 9 December 2011) (DOI:10.1002/\linebreak[0]nme.3211).

Chapter~\ref{chap:partI.virtualsurgery} is a version (with minor revisions) of ``Local Flaps: A Real-Time Finite Element Based Solution to the Plastic Surgery Defect Puzzle'' by Eftychios Sifakis, Jeffrey Hellrung, Joseph Teran (PI), Aaron Oliker, and Court Cutting,~M.D. in \textit{Studies in Health Technology and Informatics} (142:313--318, 2009) (PMID:19377176).

Chapter~\ref{chap:partI.fractureanimation} is a version (with minor revisions) of ``Geometric fracture modeling in BOLT'' by Jeffrey Hellrung (PI), Andrew Selle, Arthur Shek, Eftychios Sifakis, and Joseph Teran in \textit{SIGGRAPH 2009: Talks} (SIGGRAPH '09, pp. 7:1--7:1, New York, NY, USA, 2009) \linebreak[4] (DOI:10.1145/1597990.1597997). This research was supported by Walt Disney Animation Studios.

Chapter~\ref{chap:partII.poisson} is a version (with moderate revisions) of ``A Second Order Virtual Node Method for Elliptic Problems with Interfaces and Irregular Domains in Three Dimensions'' by Jeffrey Lee Hellrung,~Jr., Luming Wang, Eftychios Sifakis, and Joseph M. Teran (PI) in \textit{Journal of Computational Physics} (231(4):2015--2048, February 2012) (DOI:10.1016/j.jcp.2011.11.023). I wish to thank Jacob Bedrossian and James H. von Brecht for their helpful discussions; and Russell Howes and Alexey Stomakhin for submitting typographical errors and providing comments on the later drafts of the publication submission.

Chapter~\ref{chap:partII.LE} is a version (with moderate revisions) of ``A second-order virtual node algorithm for nearly incompressible linear elasticity in irregular domains'' by Yongning Zhu, Yuting Wang, Jeffrey Hellrung, Alejandro Cantarero, Eftychios Sifakis, and Joseph M. Teran (PI) (submitted for publication, 2011).
}

\vitaitem{2003, summer}{Research Student for Associate Professor Jon T. Jacobsen of Harvey Mudd College (Claremont, CA)}
\vitaitem{2004, summer}{Summer Intern at Auditude$^{\text{TM}}$, Inc. (Los Angeles, CA)}
\vitaitem{2004 - 2005}{Clinic Project Manager serving Hewlett-Packard Company (Palo Alto, CA) via Harvey Mudd College Mathematics Clinic}
\vitaitem{2005, May}{Received degree of Bachelor of Science in Mathematics (with High Distinction, Honors in Mathematics) from Harvey Mudd College}
\vitaitem{2005 - 2012}{Teaching Fellow for the Department of Mathematics, University of California Los Angeles}
\vitaitem{2005, summer}{Prof. NE MTS Level 1 at The Aerospace Corporation (El Segundo, CA)}
\vitaitem{2006, June}{Received degree of Master of Arts in Mathematics from University of California Los Angeles}
\vitaitem{2006, summer}{Prof. NE MTS Level 1 at The Aerospace Corporation}
\vitaitem{2007, summer}{Prof. NE MTS Level 1 at The Aerospace Corporation}
\vitaitem{2008, summer}{Graduate Associate at Walt Disney Animation Studios (Burbank, CA)}
\vitaitem{2010, summer}{Teaching Assistant at Park City Mathematics Institute (Park City, UT)}

\publication{M.~Hecht, D.~Buettner, \textbf{J.~Hellrung}. ``Risk assessment of real time digital control systems.'' \textit{Proceedings of the RAMS '06. Annual Reliability and Maintainability Symposium, 2006}, pp. 409--415, 2006. DOI:10.1109/RAMS.2006.1677409}
\publication{Eftychios Sifakis, \textbf{Jeffrey Hellrung}, Joseph Teran, Aaron Oliker, Court Cutting,~M.D. ``Local Flaps: A Real-Time Finite Element Based Solution to the Plastic Surgery Defect Puzzle.'' \textit{Studies in Health Technology and Informatics}, 142:313--318, 2009. PMID:19377176}
\publication{\textbf{Jeffrey Hellrung}, Andrew Selle, Arthur Shek, Eftychios Sifakis, Joseph Teran. ``Geometric fracture modeling in BOLT.'' \textit{SIGGRAPH 2009: Talks}, SIGGRAPH '09, pp. 7:1--7:1, New York, NY, USA, 2009. DOI:10.1145/1597990.1597997}
\publication{Casey L. Richardson, Jan Hegemann, Eftychios Sifakis, \textbf{Jeffrey Hellrung}, Joseph M. Teran. ``An XFEM method for modeling geometrically elaborate crack propagation in brittle materials.'' \textit{International Journal for Numerical Methods in Engineering}, 88(10):1042--1065, December 2011. DOI:10.1002/nme.3211}
\publication{\textbf{Jeffrey Lee Hellrung,~Jr.}, Luming Wang, Eftychios Sifakis, Joseph M. Teran. ``A Second Order Virtual Node Method for Elliptic Problems with Interfaces and Irregular Domains in Three Dimensions.'' \textit{Journal of Computational Physics}, 231(4): 2015--2048, February 2012. DOI:10.1016/j.jcp.2011.11.023}
\publication{Yongning Zhu, Yuting Wang, \textbf{Jeffrey Hellrung}, Alejandro Cantarero, Eftychios Sifakis, Joseph M. Teran. ``A second-order virtual node algorithm for nearly incompressible linear elasticity in irregular domains.'' (submitted for publication, 2011)}

%\presentation{}

\award{William Lowell Putnam Mathematical Competition - Top-500 Individual Placement (2001, 2004); Top-200 Individual Placement (2002, 2003); 11$^{th}$ Team Placement (2004)}
\award{Courtney S. Coleman Prize in Mathematics (awarded by HMC; 2003)}
\award{ACM (Association for Computing Machinery) Programming Contest - 20$^{th}$ (of 59) place (2003); 7$^{th}$ (of 63) place (2004)}
\award{MCM (Mathematical Contest in Modeling) - Meritorious Winner (2004)}
\award{Microsoft Imagine Cup Algorithm Invitational - 18$^{th}$ place internationally (2004)}
\award{Stavros Busenberg Prize in Applied Mathematics (awarded by HMC; 2004)}
\award{Robert Borrelli Clinic Prize for Most Outstanding Clinic Team (awarded by HMC; 2005)}
\award{Chancellor's Prize (awarded by UCLA; 2005 - 2006)}
\award{VIGRE Fellowship (awarded by UCLA Department of Mathematics; 2005 - 2009)}
\award{ICFP (International Conference in Functional Programming) Programming Contest - 80$^{th}$ (of 215+) place (2010); 95$^{th}$ (of 199) place (2011)}
\award{Google Games Santa Monica - 3$^{rd}$ place (2011)}

\abstract{
We present a collection of embedded methods to solve a variety of scientific computing problems in both $2$ and $3$ dimensions. Embedded methods make use of a structured background mesh which does not conform to the irregular geometry of the problem, such as the domain boundary. Instead, the irregular geometry is embedded within the structured mesh's elements, providing a framework to solve many problems involving crack propagation, progressive fracturing, dynamic interfaces, and shape optimization.

In Part I, we apply the mesh cutting algorithm of Sifakis et al. \cite{Sifakis07} to investigate the modeling of cracks, surgical incisions, and shattering. Specifically, we present a geometrically flexible and straightforward crack propagation method which combines the \emph{eXtended Finite Element Method} (XFEM) with \cite{Sifakis07} and an innovative integration scheme which makes use of a subordinate quadrature mesh. We also discuss the application of \cite{Sifakis07} and other advances in numerical methods to address the challenges of a virtual surgery simulator. We conclude Part I by describing a system to facilitate the modeling of cracked and shattered objects in the context of visual effects and computer animation.

In Part II, we present a numerical method utilizing virtual degrees of freedom to efficiently solve elliptic partial differential equations (specifically: Poisson's equation with interfacial jump conditions; and linear elasticity in the nearly incompressible regime) on irregular domains within a regular background Cartesian grid. Our method enforces Dirichlet boundary conditions and interfacial jump conditions weakly, formulating our system as a constrained minimization problem. In this context, we describe an algorithm to generate an associated discrete Lagrange multiplier space that allows one to derive an equivalent symmetric positive definite linear system. We provide a family of multigrid algorithms to solve this linear system with near optimal efficiency. Our method is second order accurate in $L^{\infty}$ and possesses a feature set rarely found among the broad class of embedded methods for elliptic problems.
}
